---
title: "Supplementary Code"
subtitle: "Research Article Presentation — Medicaid Expansion and Access to Health Care"
author: "Zachary Johnigan and Katia"
date: "Summer 2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Project Overview

This R Markdown file contains supplementary code for our Research Article Presentation in **Data for Policy Analysis and Management**.

## Research Article

**Sommers, Benjamin D., Robert J. Blendon, E. John Orav, and Arnold M. Epstein.**  
“Changes in Utilization and Health Among Low-Income Adults After Medicaid Expansion or Expanded Private Insurance.”  
*JAMA Internal Medicine* 176, no. 10 (2016): 1501–1509.

## Research Question

How do two alternative state approaches to expanding health insurance—traditional Medicaid expansion and Arkansas’s private insurance expansion (“private option”)—affect health care utilization and health?

## Study States

- **Kentucky:** Traditional Medicaid expansion
- **Arkansas:** Medicaid-funded private insurance expansion (“private option”)
- **Texas:** No Medicaid expansion; comparison state

## Study Period

- **2013:** Before expansion
- **2014:** First post-expansion year
- **2015:** Second post-expansion year

# Quantitative Methodology

The authors used a **difference-in-differences research design** implemented with **multivariable linear regression**.

The first analysis compared changes in Arkansas and Kentucky with changes in Texas to estimate the overall impact of coverage expansion.

The second analysis added interaction terms to compare Arkansas’s private option with Kentucky’s traditional Medicaid expansion.

# Purpose of This Supplementary Code

The original respondent-level dataset and statistical code do not appear to be included with the article. The code in this document is therefore a **teaching demonstration** of the statistical logic described in the paper rather than an exact replication of the authors’ analysis.

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Create Demonstration Dataset

```{r}
# Demonstration dataset will be created here.
```

# Inspect the Data

```{r}
# Code for examining the demonstration dataset will go here.
```

# Difference-in-Differences

## Expansion States vs. Texas

```{r}
# Simplified difference-in-differences model will go here.
```

# Multivariable Linear Regression

```{r}
# Simplified teaching version of the regression described in the article
# will go here.
```

# Arkansas Private Option vs. Kentucky Medicaid Expansion

```{r}
# Second analysis comparing the two expansion approaches will go here.
```

# Interpret the Results

This section will translate the model results into plain English and connect them to the methodology and findings reported in the article.

# Notes

- This is **supplementary teaching code**, not the authors’ original statistical code.
- The demonstration dataset is not the original survey dataset.
- The goal is to understand how the quantitative methodology described in the article works in R.
