---
title: "Supplementary Code"
subtitle: "Research Article Presentation — Medicaid Expansion and Access to Health Care"
author: "Zachary Johnigan and Katia"
date: "Summer 2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Project Overview

This R Markdown file contains supplementary code for our Research Article Presentation in **Data for Policy Analysis and Management**.

## Research Article

**Sommers, Benjamin D., Robert J. Blendon, E. John Orav, and Arnold M. Epstein.**  
“Changes in Utilization and Health Among Low-Income Adults After Medicaid Expansion or Expanded Private Insurance.”  
*JAMA Internal Medicine* 176, no. 10 (2016): 1501–1509.

## Research Question

How do two alternative state approaches to expanding health insurance—traditional Medicaid expansion and Arkansas’s private insurance expansion (“private option”)—affect health care utilization and health?

## Study States

- **Kentucky:** Traditional Medicaid expansion
- **Arkansas:** Medicaid-funded private insurance expansion (“private option”)
- **Texas:** No Medicaid expansion; comparison state

## Study Period

- **2013:** Before expansion
- **2014:** First post-expansion year
- **2015:** Second post-expansion year

# Quantitative Methodology

The authors used a **difference-in-differences research design** implemented with **multivariable linear regression**.

The first analysis compared changes in Arkansas and Kentucky with changes in Texas to estimate the overall impact of coverage expansion.

The second analysis added interaction terms to compare Arkansas’s private option with Kentucky’s traditional Medicaid expansion.

# Purpose of This Supplementary Code

The original respondent-level dataset and statistical code do not appear to be included with the article. The code in this document is therefore a **teaching demonstration** of the statistical logic described in the paper rather than an exact replication of the authors’ analysis.

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Create Demonstration Dataset

```{r}
# Create a small demonstration dataset
medicaid_demo <- data.frame(
  
  # State where each respondent lives
  state = rep(c("Arkansas", "Kentucky", "Texas"), each = 10),
  
  # Survey year
  year = rep(c(2013, 2014, 2015, 2013, 2014), 6),
  
  # Whether the state expanded coverage
  expansion = rep(c("Expansion", "Expansion", "No expansion"), each = 10),
  
  # Whether respondent reports having a personal doctor
  # 0 = No, 1 = Yes
  personal_doctor = c(
    0,0,0,1,1,1,1,1,1,1,       # Arkansas
    0,0,1,0,1,1,1,1,1,1,       # Kentucky
    0,0,1,1,0,0,1,0,1,1        # Texas
  ),
  
  # Demographic characteristics
  age = c(
    24,31,45,52,38,27,61,43,35,56,
    29,47,33,59,41,22,50,36,63,44,
    26,39,55,32,48,21,60,37,42,53
  ),
  
  sex = rep(c("Female", "Male"), 15),
  
  race_ethnicity = rep(
    c("White", "Black", "Hispanic", "White", "Black"), 6
  ),
  
  marital_status = rep(
    c("Single", "Married", "Single", "Married", "Single"), 6
  ),
  
  family_size = rep(c(1, 2, 3, 4, 2), 6),
  
  education = rep(
    c("High school", "Some college", "College",
      "High school", "Some college"), 6
  ),
  
  # Approximate annual income for demonstration purposes
  income = c(
    12000,18000,15000,22000,14000,19000,11000,21000,16000,13000,
    17000,12500,20000,14500,23000,11500,18500,15500,21500,13500,
    16000,20500,12000,17500,22500,13000,19500,15000,21000,14000
  ),
  
  residence = rep(c("Urban", "Rural"), 15)
)

# View the demonstration dataset
medicaid_demo

# Optional: save the demonstration dataset
# write_csv(medicaid_demo, "medicaid_demo.csv")
```

# Inspect the Data

```{r}
# Look at the demonstration dataset
View(medicaid_demo)

# Check the structure and variable types
str(medicaid_demo)

# Summarize all variables
summary(medicaid_demo)

# Count respondents by state
table(medicaid_demo$state)

# Count respondents by year
table(medicaid_demo$year)

# Count respondents by expansion status
table(medicaid_demo$expansion)

# Count who has a personal doctor
# 0 = No, 1 = Yes
table(medicaid_demo$personal_doctor)

# Examine demographic variables
table(medicaid_demo$sex)
table(medicaid_demo$race_ethnicity)
table(medicaid_demo$marital_status)
table(medicaid_demo$education)
table(medicaid_demo$residence)

# Summarize numeric demographic variables
summary(medicaid_demo$age)
summary(medicaid_demo$family_size)
summary(medicaid_demo$income)
```

# Difference-in-Differences

## Expansion States vs. Texas

```{r}
# Simplified difference-in-differences model
# Compare Arkansas and Kentucky (expansion) with Texas (no expansion)

did_model <- lm(
  personal_doctor ~ expansion * factor(year),
  data = medicaid_demo
)

# View the results
summary(did_model)
```

# Multivariable Linear Regression

```{r}
# Multivariable linear regression
# Same basic comparison as the DiD model, but now we also account
# for characteristics of the respondents

multivariable_model <- lm(
  personal_doctor ~ expansion * factor(year) +
    age +
    sex +
    race_ethnicity +
    marital_status +
    family_size +
    education +
    income +
    residence,
  data = medicaid_demo
)

# View the results
summary(multivariable_model)
```

# Arkansas Private Option vs. Kentucky Medicaid Expansion

```{r}
# Compare Arkansas's private option with Kentucky's
# traditional Medicaid expansion

expansion_comparison <- medicaid_demo %>%
  filter(state %in% c("Arkansas", "Kentucky"))

# Run the comparison model
private_vs_medicaid_model <- lm(
  personal_doctor ~ state * factor(year) +
    age +
    sex +
    race_ethnicity +
    marital_status +
    family_size +
    education +
    income +
    residence,
  data = expansion_comparison
)

# View the results
summary(private_vs_medicaid_model)
```
# Save the complete model results to a text file
```{r}
capture.output(
  summary(private_vs_medicaid_model),
  file = "private_vs_medicaid_results.txt"
)
```

# Interpret the Results
Our demonstration model compares changes in access to a personal doctor between Arkansas and Kentucky over time. Arkansas represents the private insurance expansion approach, while Kentucky represents traditional Medicaid expansion.

The most important results for this comparison are the interaction terms between state and year. These tell us whether Kentucky's change in access differed from Arkansas's change during the post-expansion years.

For 2014, the interaction estimate was 0.233 with a p-value of 0.704. For 2015, the interaction estimate was 0.957 with a p-value of 0.233. Because both p-values are greater than 0.05, our small demonstration dataset does not provide statistically significant evidence that changes in access to a personal doctor differed between Arkansas and Kentucky.

These results should not be interpreted as reproducing the findings of the actual study. Our dataset is a small, artificial dataset created only to demonstrate how the authors' statistical approach works. The actual researchers used a much larger sample and a more sophisticated analysis.

# AI-Use Disclosure
AI was used to assist in the creation of this simplified replica model for educational purposes. The model is in no way a viable substitute for the authors' actual dataset, statistical analysis, or results. Instead, it serves as a learning tool for imagining and better understanding how the authors may have conducted the linear regression portion of the study described in the Statistical Analysis section of the Methods.

AI assistance was used to help create the small demonstration dataset, translate the statistical methodology described in the article into simplified R code, troubleshoot the code, and interpret the resulting model output in plain English.

The demonstration data are artificial, and the results produced by the replica model should not be interpreted as reproducing or validating the findings of the original study. Any conclusions about the study itself are based on the authors' published research rather than the AI-assisted demonstration.



- This is **supplementary teaching code**, not the authors’ original statistical code.
- The demonstration dataset is not the original survey dataset.
- The goal is to understand how the quantitative methodology described in the article works in R.
