---
title: "Final Project Presentation"
output: html_document
---
#------------------------------------------------------------------------------#
# This file prepares Mass Shootings data for the Final Project Presentation of the Data for Policy Analysis class.
# PROJECT NAME : Data for Policy Analysis, Final Project Presentation
# DATA SETS USED BY THIS CODE : mass_shootings.csv
# R VERSION : 4.5.3
# AUTHOR : Zachary Johnigan
# DATE CREATED : 08-13-2026
# NOTES : 
#------------------------------------------------------------------------------#

#------------------------------------------------------------------------------#
# 1. Set up working space and import packages
#------------------------------------------------------------------------------#

```{r}
rm(list=ls())
options("error") 
options(lifecycle_disable_verbose_retirement = TRUE)

```

# Import packages

```{r}
package.list <- c("tidyverse", "readr")

for (p in package.list){
  if (!p %in% installed.packages()[, "Package"]) install.packages(p)
  library(p, character.only=TRUE)
}

```

# Set up working directory 

## check your directory
```{r}
getwd()

```

## set up working directory 

```{r}
my_directory   <- "/home/zacharyjohnigan/UChicago_Student/2_Data_For_Policy_Analys_-_Mgmt/Final_Project_Presentation/"
my_data        <- paste0(my_directory, "Data/")
output_dir     <- paste0(my_directory, "Presentation")

```

#------------------------------------------------------------------------------#
# 2. Load dataset
#------------------------------------------------------------------------------#

# Import Mass Shootings dataset
## To import .CSV files use this code
## data <- read.csv("[full directory here]")
```{r}
#load(paste0(my_data, "mass_shootings.csv"))
data <- read.csv("~/UChicago_Student/2_Data_For_Policy_Analys_-_Mgmt/Final_Project_Presentation/Data/mass_shootings.csv")

```

#------------------------------------------------------------------------------#
# 3. Basic exploratory data inspection
#------------------------------------------------------------------------------#

# Understand variable class for location and location_type
```{r, eval=FALSE}
# This chunk examines the class and summary of each variable to better understand the dataset before beginning any analysis.
class(data$location) #character
summary(data$location) #nominal or discrete?

class(data$location_type) #character
summary(data$location_type) #nominal or discrete?

glimpse(data)

```

#------------------------------------------------------------------------------#
# 4. Basic cleaning and manipulation
#------------------------------------------------------------------------------#

# Rename dataset
```{r}
mass_shootings_data <- data

```

# Subset dataset and rename variables
```{r}
nsece_2019_wf_subset <- 
  nsece_2019_wf %>%
  select(WF9_PROFDEV_WRKSHP, 
         WF9_PROFDEV_COACH, 
         WF9_PROFDEV_MEETING, 
         WF9_PROFDEV_COURSE,
         WF9_WORK_WAGE) %>%
  rename(part_workshop = WF9_PROFDEV_WRKSHP,
         part_coaching = WF9_PROFDEV_COACH, 
         part_meeting = WF9_PROFDEV_MEETING, 
         part_course = WF9_PROFDEV_COURSE, 
         hr_wage = WF9_WORK_WAGE) 

```

# I suggest keeping the factor variables for ordinal variables but work with numeric / character values for all other types of variables.Since all the predictor variables are nominal, I will convert them as character variables.

# Recode professional development variables
```{r}
nsece_2019_wf_recode <- 
  nsece_2019_wf_subset %>%
  mutate(part_workshop_char = as.character(part_workshop), # use mutate() to create new character variables while keeping the original variables unchanged
         part_workshop_char = case_when(
           part_workshop == "(1) Yes" ~ "Yes", # recode the variable responses to simpler labels
           part_workshop == "(2) No" ~ "No"),
           
         part_coaching_char = as.character(part_coaching),
         part_coaching_char = case_when(
           part_coaching == "(1) Yes" ~ "Yes",
           part_coaching == "(2) No" ~ "No"), 
         
         part_meeting_char = as.character(part_meeting),
         part_meeting_char = case_when(
           part_meeting == "(1) Yes" ~ "Yes",
           part_meeting == "(2) No" ~ "No"), 
         
         part_course_char = as.character(part_course),
         part_course_char = case_when(
           part_course == "(1) Yes" ~ "Yes",
           part_course == "(2) No" ~ "No"))
         
```
This chunk creates a new dataset with cleaner versions of several variables. It uses mutate() to add new columns and case_when() to replace coded responses like “(1) Yes” and “(2) No” with the simpler labels “Yes” and “No”. The original variables are preserved, while the new _char variables make the data easier to analyze, graph, and interpret.

# Create new professional development variable
```{r}
nsece_2019_wf_pd <- 
  nsece_2019_wf_recode %>%
  mutate(pd = ifelse( # use mutate() to create a new variable called pd
    part_workshop_char == "Yes" | part_coaching_char == "Yes" | part_meeting_char == "Yes" | part_course_char == "Yes", 
    1, 0)) # If the respondent answered "Yes" to at least one of the four activities (workshop, coaching, meeting, or course), code pd as 1.
```

# Otherwise, code pd as 0.

## Let's examine the new variable created
```{r}
class(nsece_2019_wf_pd$pd) #numeric
table(nsece_2019_wf_pd$pd) #the variable was created as expected 
summary(nsece_2019_wf_pd$pd) 
glimpse(nsece_2019_wf_pd)
```

# About the variables
```{r, eval=FALSE}
part_workshop = participated_in_workshop,
# Indicates whether the respondent participated in a professional development workshop.
part_coaching = participated_in_coaching,
# Indicates whether the respondent participated in professional coaching.
part_meeting = participated_in_professional_meeting,
# Indicates whether the respondent attended a professional meeting.
part_course = completed_professional_course,
# Indicates whether the respondent completed a professional development course.
hr_wage = hourly_wage,
# The respondent's hourly wage.
part_workshop_char = participated_in_workshop_char,
# A "Yes"/"No" version of the workshop participation variable.
part_coaching_char = participated_in_coaching_char,
# A "Yes"/"No" version of the coaching participation variable.
part_meeting_char = participated_in_professional_meeting_char,
# A "Yes"/"No" version of the meeting participation variable.
part_course_char = completed_professional_course_char,
# A "Yes"/"No" version of the course participation variable.
pd = participated_in_any_professional_development
# Indicates whether the respondent participated in at least one professional development activity (1 = Yes, 0 = No, NA = Missing).
```
