---
title: "Final Project Presentation"
author: "Zachary Johnigan"
output: html_document
---

# Research Question

**Are differences in mass-shooting location type associated with differences in the number of fatalities per incident?**

# Class and teacher comments
- Perhaps look into grouping by region
- Find the code book for your dataset

# Clean Environment
```{r}
rm(list=ls())
options("error")
options(lifecycle_disable_verbose_retirement = TRUE)
```

# Load Packages
```{r}
library(tidyverse)
library(rcis)
```

# Load Dataset
```{r}
data("mass_shootings")
```

# Create Working Dataset
```{r}
mass_shootings_project <- mass_shootings
```

# Inspect the Dataset
```{r}
glimpse(mass_shootings_project)
view(mass_shootings)
names(mass_shootings)
```

```{r}
names(mass_shootings_project)
```

```{r}
head(mass_shootings_project)
```

# Variables of Interest
For this analysis:
- **Independent variable:** `location_type`
- **Dependent variable:** `fatalities`

```{r}
class(mass_shootings_project$location_type)
class(mass_shootings_project$fatalities)
```

# Location Type  

## Frequencies and Percentages
```{r}
location_table <-
  mass_shootings_project %>%
  count(location_type) %>%
  mutate(percent = (n / sum(n)) * 100)

location_table
```
This code counts how many mass shootings occurred in each type of location and calculates what percentage of all shootings each location type represents.

## Bar Graph
```{r}
ggplot(mass_shootings_project,
       aes(x = location_type)) +
  geom_bar() +
  coord_flip() +
  labs(
    title = "Mass Shootings by Location Type",
    x = "Location Type",
    y = "Number of Incidents"
  )
```
This code creates a horizontal bar graph showing how many mass-shooting incidents occurred in each location type, making the categories easy to compare visually.

# Fatalities

## Descriptive Statistics

```{r}
mass_shootings_project %>%
  summarise(
    mean = mean(fatalities, na.rm = TRUE),
    median = median(fatalities, na.rm = TRUE),
    sd = sd(fatalities, na.rm = TRUE),
    min = min(fatalities, na.rm = TRUE),
    max = max(fatalities, na.rm = TRUE),
    IQR = IQR(fatalities, na.rm = TRUE)
  )
```

## Histogram

```{r}
ggplot(mass_shootings_project,
       aes(x = fatalities)) +
  geom_histogram() +
  labs(
    title = "Distribution of Fatalities",
    x = "Number of Fatalities",
    y = "Frequency"
  )
```

# Compare Location Type and Fatalities

## Group Means

```{r}
mass_shootings_project %>%
  group_by(location_type) %>%
  summarise(
    mean_fatalities = mean(fatalities, na.rm = TRUE)
  )
```

## More Detailed Group Comparison

```{r}
mass_shootings_project %>%
  group_by(location_type) %>%
  summarise(
    n = n(),
    mean_fatalities = mean(fatalities, na.rm = TRUE),
    median_fatalities = median(fatalities, na.rm = TRUE),
    sd_fatalities = sd(fatalities, na.rm = TRUE)
  )
```

# Boxplot

```{r}
ggplot(mass_shootings_project,
       aes(x = location_type,
           y = fatalities)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Fatalities by Location Type",
    x = "Location Type",
    y = "Number of Fatalities"
  )
```

# OLS Regression

```{r}
model <-
  lm(
    fatalities ~ location_type,
    data = mass_shootings_project
  )

summary(model)
```

# Regression Results

```{r}
summary(model)$coefficients
```

# Model Fit

```{r}
summary(model)$r.squared
summary(model)$adj.r.squared
```

# Regression Diagnostics

## Residual Plot

```{r}
plot(
  fitted(model),
  resid(model)
)

abline(h = 0)
```

## Histogram of Residuals

```{r}
hist(resid(model))
```

# Interpretation

**Research question:**  
Are differences in mass-shooting location type associated with differences in the number of fatalities per incident?

**Independent variable:**  
Location type

**Dependent variable:**  
Number of fatalities

**Main finding:**  


**Policy relevance:**  

