---
title: "ProblemSet_1"
author: "Zachary Johnigan"
date: "2026-07-22"
output: pdf_document
---

Zachary Johnigan  
Aida Pacheco-Applegate  
Data For Policy Analys/Mgmt SSAD 48500/1 [60558]  
Due: Sun Aug 2, 2026 11:59pm  

\centering
ProblemSet_1

1. Prepare the data for analysis. Subset the NSECE dataset to include the following
variables:
o WF9_WORK_WAGE (hourly wage)
o WF9_WORK_ROLE (staff role)
o WF9_WORK_YRS (number of years worked in the program)
o WF9_CHAR_HISP (Hispanic/Latino ethnicity)
o WF9_CHAR_RACE (race)
o WF9_CHAR_EDUC (highest level of education completed)
o WF9_CHAR_GENDER (gender)

- Rename the variables using clear, descriptive names. 
- Perform any data cleaning you consider necessary. 
- For each variable, identify its level of measurement (nominal, ordinal, or 
interval/ratio) and whether it is discrete or continuous. 

Include the appropriate descriptive statistics, tables, or graphs, as well as 
measures of central tendency, dispersion, and shape when applicable. 

Describe and interpret your findings.
(4 points)
```{r}
```

2. Examine the following relationships. For each analysis, include the appropriate
statistical test to determine whether the observed differences or associations are
statistically significant. Interpret your results in the context of the ECE workforce. (4
points)
o a. Hourly wage and staff role
o b. Hourly wage and number of years worked in the program
o c. Hourly wage and race/ethnicity
o d. Hourly wage and educational attainment
o e. Hourly wage and gender
```{r}
```

3. Policy implications. Based on your analyses, describe the segment(s) of the ECE
workforce that appear to be most affected by low wages. What characteristics would
you consider when designing a policy to improve compensation in the ECE sector?
Support your answer using evidence from your analyses. (2 points)
```{r}
```

