---
title: "ProblemSet_1"
author: "Zachary Johnigan"
date: "2026-07-22"
output: pdf_document
---

Zachary Johnigan  
Aida Pacheco-Applegate  
Data For Policy Analys/Mgmt SSAD 48500/1 [60558]  
Due: Sun Aug 2, 2026 11:59pm  

\centering
ProblemSet_1

1. Prepare the data for analysis. Subset the NSECE dataset to include the following variables:
o WF9_WoRK_WAGE (hourly wage)
o WF9_WoRK_RoLE (staff role)
o WF9_WoRK_YRS (number of years worked in the program)
o WF9_CHAR_HISP (Hispanic/Latino ethnicity)
o WF9_CHAR_RACE (race)
o WF9_CHAR_EDUC (highest level of education completed)
o WF9_CHAR_GENDER (gender)

[x] Rename the variables using clear, descriptive names. 
[-] Perform any data cleaning you consider necessary. 
[-] For each variable, identify its level of measurement (nominal, ordinal, or 
interval/ratio) and whether it is discrete or continuous. 

Include the appropriate descriptive statistics, tables, or graphs, as well as measures of central tendency, dispersion, and shape when applicable. 

Describe and interpret your findings.
(4 points)

# 1. Set up working space and import packages
```{r}
# This cleans up everything in the "Environment" pane.
rm(list=ls())
options("error") 
options(lifecycle_disable_verbose_retirement = TRUE)
```

```{r}
# Import packages
package.list <- c("tidyverse")

# read the `package.list` for the list of packages we will be needing
# Check what packages are installed already 
# If any of the packages in the c() list are not currently installed, install them.
# Note: Check the "Packages" tab to verify that they were installed. If they were not, then check the box and click "Install".
for (p in package.list){
  if (!p %in% installed.packages()[, "Package"]) install.packages(p)
  library(p, character.only=TRUE)
}
```

```{r}
# Set up working directory 

## check your directory
getwd()

## set up working directory 
my_directory   <- "/home/zacharyjohnigan/UChicago_Student/2_Data_For_Policy_Analys_-_Mgmt/"
my_directory

my_data        <- paste0(my_directory, "Data/")
my_data

output_dir     <- paste0(my_directory, "Problem_Sets/Problem_Set_1/")
output_dir
```

# 2. Load dataset
```{r}
# Import NSECE 2019 workforce questionnnaire (since the data is already in R
# format we just need to use the line 52 of this code to load it into our work 
# environment)

load(paste0(my_data, "37941-0005-Data.rda"))
```

# 4. Basic cleaning of the data
```{r}
# Rename dataset
nsece_2019_wf <- da37941.0005
```

# Manipulation if the data
# Subset the NSECE dataset to include the following variables:
```{r}
# WF9_WORK_WAGE (hourly wage)
# WF9_WORK_ROLE (staff role)
# WF9_WORK_YRS (number of years worked in the program)
# WF9_CHAR_HISP (Hispanic/Latino ethnicity)
# WF9_CHAR_RACE (race)
# WF9_CHAR_EDUC (highest level of education completed)
# WF9_CHAR_GENDER (gender)

hourly_wage
staff_role
number_of_years_worked_in_the_program
Hispanic_Latino_ethnicity
race
highest_level_of_education_completed
gender
```

# Subset the NSECE dataset
```{r}
nsece_2019_wf_subset <- 
  nsece_2019_wf %>%
  select(
    WF9_WORK_WAGE, 
    WF9_WORK_ROLE, 
    WF9_WORK_YRS, 
    WF9_CHAR_HISP,
    WF9_CHAR_RACE,
    WF9_CHAR_EDUC,
    WF9_CHAR_GENDER) %>%
  rename(
    hourly_wage = WF9_WORK_WAGE, 
    staff_role = WF9_WORK_ROLE, 
    number_of_years_worked_in_the_program = WF9_WORK_YRS, 
    Hispanic_Latino_ethnicity = WF9_CHAR_HISP,
    race = WF9_CHAR_RACE,
    highest_level_of_education_completed = WF9_CHAR_EDUC,
    gender = WF9_CHAR_GENDER
  )
```


# Each variable's level of measurement (nominal, ordinal, or interval/ratio) and whether it is discrete or continuous.
According to the 37941-0005-User_guide.pdf, "manual provides information on the 2019 National Survey of Early Care and
Education (NSECE) workforce public-use data files"(), and the lecture we can define each of the 7 variables as follows.

WF9_WORK_WAGE - hourly_wage - Hourly wage for Workforce respondent (top-coded)
Level of Measurement: Ratio, Continuous

WF9_WORK_ROLE - staff_role - Role of Workforce respondent
Level of Measurement: Ordinal, Discreate

WF9_WORK_YRS - number_of_years_worked_in_the_program - Num of years respondent worked in program
Level of Measurement: Ratio, Continuous

WF9_CHAR_HISP - Hispanic_Latino_ethnicity - Respondent is of Hispanic/Latino descent
Level of Measurement: Nominal, Discrete

WF9_CHAR_RACE - race - Respondent's race
Level of Measurement: Nominal, Discrete

WF9_CHAR_EDUC - highest_level_of_education_completed - Highest education level obtained
Level of Measurement: Ordinal, Discrete

WF9_CHAR_GENDER - gender - The respondent’s self-reported gender
Level of Measurement: Nominal, Discrete

# Citations
NSECE Project Team (National Opinion Research Center). (2019). National Survey of Early Care and Education (NSECE), [United States], 2019: NSECE user’s guide for workforce public-use data file (ICPSR 37941). Inter-university Consortium for Political and Social Research.

Pacheco-Applegate, A. (2026). Data for Policy Analysis and Management: Class 2—Data measurement and variables [PowerPoint slides]. Harris School of Public Policy, University of Chicago.
















2. Examine the following relationships. For each analysis, include the appropriate
statistical test to determine whether the observed differences or associations are
statistically significant. Interpret your results in the context of the ECE workforce. (4
points)
o a. Hourly wage and staff role
o b. Hourly wage and number of years worked in the program
o c. Hourly wage and race/ethnicity
o d. Hourly wage and educational attainment
o e. Hourly wage and gender
```{r}
```






















































3. Policy implications. Based on your analyses, describe the segment(s) of the ECE
workforce that appear to be most affected by low wages. What characteristics would
you consider when designing a policy to improve compensation in the ECE sector?
Support your answer using evidence from your analyses. (2 points)
```{r}
```

