---
title: "Methods We Have Covered in Class"
author: "Zachary Johnigan"
output: html_document
---

# Methods We Have Covered in Class

The following code chunks are placeholders for methods that may be useful as the project develops. Not all methods will necessarily be used in the final analysis.

# Clean Environment
```{r}

```

## Load Packages
```{r}

```

## Load Built-In Dataset
```{r}

```

## Create Working Dataset
```{r}
cars 
```

## Import Dataset
```{r}

```

## Inspect the Dataset
```{r}

```

## Data Cleaning and Recoding

```{r}
# Data cleaning and recoding

```

## Variable Types and Levels of Measurement

```{r}
# Examine variable types and classes

```

## Descriptive Statistics

```{r}
# Mean, median, mode, range, standard deviation, and IQR

```

## Frequencies and Percentages

```{r}
# Frequency and percentage tables

```

## Cross-Tabulations

```{r}
# Examine frequencies across two categorical variables

```

## Bar Graph

```{r}
# Bar graph

```

## Histogram

```{r}
# Histogram

```

## Boxplot

```{r}
# Boxplot

```

## Scatterplot

```{r}
# Scatterplot

```

## Comparing Groups

```{r}
# Compare groups

```

## Binary Variables

```{r}
# Create binary/dummy variables

# Reference group:

```

## Correlation

```{r}
# Examine correlation between variables

```

## Simple OLS Regression

```{r}
# Estimate a simple OLS regression model

```

## Multiple OLS Regression

```{r}
# Estimate a multiple OLS regression model

```

## Regression Results

```{r}
# Examine coefficient estimates and statistical significance

```

## Model Fit

```{r}
# Examine R-squared and adjusted R-squared

```

## Regression Diagnostics

### Residual Plot

```{r}
# Create residual plot

```

### Histogram of Residuals

```{r}
# Examine distribution of model residuals

```

# Interpretation

**Relationship being examined:**

**Independent variable(s):**

**Dependent variable:**

**Main finding:**

**Policy relevance:**
