---
title: "Methods We Have Covered in Class"
author: "Zachary Johnigan"
output: html_document
---

# Methods We Have Covered in Class

The following code chunks are placeholders for methods that may be useful as the project develops. Not all methods will necessarily be used in the final analysis.

# Clean Environment
```{r}
rm(list=ls())
options("error") 
options(lifecycle_disable_verbose_retirement = TRUE)

```

## Load Packages
```{r}
library(tidyverse)
```

## Load Built-In Dataset
```{r}
data(mtcars)
```

## Create Working Dataset
```{r}
cars <- mtcars
```

## Import Dataset
```{r}
# Use the code below if the dataset is a CSV file
#mass_shootings <- read_csv("mass_shootings.csv")
```

## Inspect the Dataset
```{r}

```

## Data Cleaning and Recoding

```{r}
# Data cleaning and recoding
cars <-
  cars %>%
  mutate(
    transmission = ifelse(am == 1, "Manual", "Automatic")
  )
```

## Variable Types and Levels of Measurement

```{r}
# Examine variable types and classes
class(cars$mpg)
class(cars$cyl)
class(cars$transmission)
```

## Descriptive Statistics

```{r}
# Descriptive statistics
cars %>%
  summarise(
    mean = mean(mpg),
    median = median(mpg),
    sd = sd(mpg),
    min = min(mpg),
    max = max(mpg),
    IQR = IQR(mpg)
  )
```

## Frequencies and Percentages

```{r}
# Frequency and percentage table
cars %>%
  count(cyl) %>%
  mutate(percent = (n / sum(n)) * 100)
```

## Cross-Tabulations

```{r}
# Cross-tabulation
table(cars$cyl, cars$transmission)
```

## Bar Graph

```{r}
# Bar graph
ggplot(cars, aes(x = factor(cyl))) +
  geom_bar()
```

## Histogram

```{r}
# Histogram
ggplot(cars, aes(x = mpg)) +
  geom_histogram()
```

## Boxplot

```{r}
# Boxplot
ggplot(cars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot()
```

## Scatterplot

```{r}
# Scatterplot
ggplot(cars, aes(x = wt, y = mpg)) +
  geom_point()
```

## Comparing Groups

```{r}
# Compare groups
cars %>%
  group_by(cyl) %>%
  summarise(mean_mpg = mean(mpg))
```

## Binary Variables

```{r}
# Create binary/dummy variable
cars <-
  cars %>%
  mutate(
    manual_d = ifelse(am == 1, 1, 0)
  )

# Reference group: Automatic
```

## Correlation

```{r}
# Correlation
cor(cars$wt, cars$mpg)
```

## Simple OLS Regression

```{r}
# Simple OLS regression
model_simple <- lm(
  mpg ~ wt,
  data = cars
)

summary(model_simple)
```

## Multiple OLS Regression

```{r}
# Multiple OLS regression
model_multiple <- lm(
  mpg ~ wt + hp,
  data = cars
)

summary(model_multiple)
```

## Regression Results

```{r}
# Regression coefficients and statistical significance
summary(model_multiple)$coefficients
```

## Model Fit

```{r}
# Model fit
summary(model_multiple)$r.squared
summary(model_multiple)$adj.r.squared
```

## Regression Diagnostics

### Residual Plot

```{r}
# Residual plot
plot(
  fitted(model_multiple),
  resid(model_multiple)
)
abline(h = 0)
```

### Histogram of Residuals

```{r}
# Histogram of residuals
hist(resid(model_multiple))
```

# Interpretation

**Relationship being examined:**

**Independent variable(s):**

**Dependent variable:**

**Main finding:**

**Policy relevance:**
